---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# DirectLiNGAM

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
<!-- badges: end -->

LiNGAM is a new method for estimating structural equation models or linear Bayesian networks. It is based on using the non-Gaussianity of the data.

This package is a port of the Python lingam package to R.

- [The LiNGAM Project](https://sites.google.com/view/sshimizu06/lingam)
- [lingam](https://github.com/cdt15/lingam)

`DirectLiNGAM` is a port to R of the [LiNGAM](https://github.com/cdt15/lingam) package (LiNGAM: Linear Non-Gaussian Acyclic Model), which is available in Python.

This is currently an alpha version under development, and we are releasing it for the purpose of testing and gathering feedback.


## Features

* Implementation of the Direct LiNGAM algorithm
* Stability assessment of causal structures using the bootstrap method
* Visualization of estimation results using DiagrammeR


## Important Notes

* This package does not include all the features of the Python version.
* This package also includes features that are not present in the Python version.


## Installation

You can install the development version of DirectLiNGAM from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("morimotoosamu/DirectLiNGAM")
```


## Requirements

- DiagrammeR


## Example

https://lingam.readthedocs.io/en/latest/tutorial/lingam.html を再現する

サンプルデータの呼び出しとDirect LiNGAMの実行。

```{r example}
library(DirectLiNGAM)
data(LiNGAM_sample_1000)

model <- direct_lingam(LiNGAM_sample_1000)
```

推定された因果順序。

```{r causal_order}
cat("Index: ", model$causal_order, "\n")
cat("Names: ", colnames(LiNGAM_sample_1000)[model$causal_order], "\n")
```

推定された隣接行列

```{r adjacency_matrix}
B_hat <- model$adjacency_matrix
colnames(B_hat) <- rownames(B_hat) <- colnames(LiNGAM_sample_1000)
cat("\n--- Estimated Adjacency Matrix ---\n")
print(round(B_hat, 3))
```


推定された因果グラフの描画

```{r plot_adjacency}
plot_adjacency_diagrammer(
  B_hat,
  threshold = 1,
  labels = colnames(LiNGAM_sample_1000),
  graph_label = "Estimated Causal Structure (Direct LiNGAM)",
  rankdir = "TB",
  shape = "ellipse",
  fillcolor = "lightyellow"
)
```


総合因果効果の推定

```{r total_effects}
LiNGAM_sample_1000 |>
  estimate_all_total_effects(model) |>
  round(3)
```



事前知識を用いた推定

x0からx5の6変数のデータセット、x3が外生変数、x1、x4、x5がシンク変数であることが事前知識。

```{r make_prior_knowledge1}
data("LiNGAM_sample_1000")

X <- LiNGAM_sample_1000

# インデックスで指定
pk1 <- make_prior_knowledge(
  n_variables         = 6,
  exogenous_variables = 4,  #外生変数x3は4個目の変数
  sink_variables = c(2, 5, 6) #シンク変数x1, x4, x5は2,5,6個目の変数
)

pk1

model_pk1 <- direct_lingam(X, prior_knowledge = pk1)

# 因果順
colnames(X)[model_pk1$causal_order]

# 影響力マトリクス
B_pk <- model_pk1$adjacency_matrix
colnames(B_pk) <- rownames(B_pk) <- colnames(X)
round(B_pk, 3)

plot_adjacency_diagrammer(
  B_pk,
    threshold = 1,
  labels      = colnames(X),
  graph_label = "Estimated (with Prior Knowledge)",
  rankdir     = "LR",
  shape       = "circle",
  fillcolor   = "lightgreen"
)
```


変数名で指定

```{r make_prior_knowledge2}
data("LiNGAM_sample_1000")

X <- LiNGAM_sample_1000

pk2 <- make_prior_knowledge(
  n_variables         = 6,
  exogenous_variables = "x3",
  sink_variables = c("x1", "x4", "x5"),
  paths = list(c("x3", "x0"), c("x3", "x2")),
  no_paths = list(c("x3", "x5")),
  labels = c("x0", "x1", "x2", "x3", "x4", "x5")
)

pk2

model_pk2 <- direct_lingam(X, prior_knowledge = pk2)

# 因果順
colnames(X)[model_pk2$causal_order]

# 影響力マトリクス
B_pk <- model_pk2$adjacency_matrix
colnames(B_pk) <- rownames(B_pk) <- colnames(X)
round(B_pk, 3)

plot_adjacency_diagrammer(
  B_pk,
  threshold = 0.1,
  labels      = colnames(X),
  graph_label = "Estimated (with Prior Knowledge)",
  rankdir     = "LR",
  shape       = "circle",
  fillcolor   = "lightgreen"
)

```

誤差変数間の独立性検定
LiNGAMの仮定が破れているかどうかを確認するために、誤差変数間の独立性のp値を算出することができる。得られた行列のi行j列にある値は、誤差変数𝑒𝑖と𝑒𝑗の独立性のp値を示す。

```{r}
# サンプルデータの呼び出し
data(LiNGAM_sample_1000)

# Direct LiNGAM の実行
result <- direct_lingam(LiNGAM_sample_1000)

# p 値の計算（デフォルト: Spearman）
p_vals <- get_error_independence_p_values(LiNGAM_sample_1000, result)
round(p_vals, 3)

# Kendall で計算
p_vals_k <- get_error_independence_p_values(LiNGAM_sample_1000, result, method = "kendall")
round(p_vals_k, 3)
```


## Licence

MIT License

Original work: Copyright (c) 2019 T.Ikeuchi, G.Haraoka, M.Ide, W.Kurebayashi, S.Shimizu

Portions of this work: Copyright (c) 2026 O.Morimoto


## Feedback

Please submit bug reports and feature requests via GitHub Issues.

