---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
set.seed(0)
```


# lingamr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
<!-- badges: end -->

LiNGAM is a method for estimating structural equation models or linear Bayesian networks. It is based on using the non-Gaussianity of the data.

`lingamr` is a port to R of the [LiNGAM](https://github.com/cdt15/lingam) package (LiNGAM: Linear Non-Gaussian Acyclic Model), which is available in Python.

- [The LiNGAM Project](https://sites.google.com/view/sshimizu06/lingam)
- [lingam (Python)](https://github.com/cdt15/lingam)

This is currently an alpha version under development, and we are releasing it for the purpose of testing and gathering feedback.


## Features

* Implementation of the Direct LiNGAM algorithm
* Stability assessment of causal structures using the bootstrap method, including causal-order stability
* Model diagnostics: residual independence / normality tests and a one-call `summary_lingam()`
* Visualization with DiagrammeR (interactive) and ggplot2 `autoplot()` (static)
* broom-style tidiers (`tidy()` / `glance()`)

This package does not include all the features of the Python version, and it also includes some features that are not present in the Python version.


## Installation

You can install the development version of `lingamr` from [GitHub](https://github.com/morimotoosamu/lingamr) with:

``` r
# install.packages("pak")
pak::pak("morimotoosamu/lingamr")
```

Some functionality relies on the following suggested packages: `DiagrammeR` (interactive plots), `igraph` and `ggplot2` (static `autoplot()` graphs and QQ plots), `glmnet` (adaptive LASSO), and `nortest` / `tseries` (residual tests).


## Quick start

```{r quickstart}
library(lingamr)

# Generate sample data from a 6-variable LiNGAM model
x <- generate_lingam_sample_6(n = 1000)

# Estimate the causal structure with Direct LiNGAM
model <- lingam_direct(x$data)

# Estimated causal order (as variable names)
colnames(x$data)[model$causal_order]
```

```{r quickstart_plot}
# Visualize the estimated causal graph
model$adjacency_matrix |>
  plot_adjacency(
    labels    = colnames(model$adjacency_matrix),
    title     = "Estimated Causal Structure (Direct LiNGAM)",
    rankdir   = "TB",
    shape     = "ellipse",
    fillcolor = "lightgreen"
  )
```


## Learn more

For a full walkthrough — prior knowledge, total causal effects, residual
independence and normality tests, and bootstrap (including parallel execution) —
see the vignette:

``` r
vignette("lingamr")
```

It is also available online at the
[package website](https://morimotoosamu.github.io/lingamr/).


## Licence

MIT License

Original work: Copyright (c) 2019 T.Ikeuchi, G.Haraoka, M.Ide, W.Kurebayashi, S.Shimizu

Portions of this work: Copyright (c) 2026 O.Morimoto


## References

### Algorithm
- Shimizu, S. et al. (2011). DirectLiNGAM: A direct method for learning
  a linear non-Gaussian structural equation model.
  *Journal of Machine Learning Research*, 12, 1225-1248.

### Original Implementation (Python)
- Ikeuchi, T. et al. (2023). Python package for LiNGAM algorithms.
  *Journal of Machine Learning Research*, 24(14), 1-7.
  https://github.com/cdt15/lingam

### Books
- 清水昌平(2017)『統計的因果探索』講談社.
- 梅津佑太・西村龍映・上田勇祐(2020)『スパース回帰分析とパターン認識』講談社.
- 鈴木譲(2025)『グラフィカルモデルと因果探索100問 with R』共立出版.

### R Packages Referenced
- G. Kikuchi (2020). rlingam
  https://github.com/gkikuchi/rlingam


## Acknowledgments

We thank the following people and organization for their support in developing this package:

- Hirohiko Asano
- Yoshiyuki Okuse
- Takahiko Umeyama
- JMRA (Japan Marketing Research Association)


## Feedback

Please submit bug reports and feature requests via GitHub Issues.
