---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# DirectLiNGAM

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
<!-- badges: end -->

LiNGAM is a new method for estimating structural equation models or linear Bayesian networks. It is based on using the non-Gaussianity of the data.

This package is a port of the Python lingam package to R.

- [The LiNGAM Project](https://sites.google.com/view/sshimizu06/lingam)
- [lingam](https://github.com/cdt15/lingam)

`DirectLiNGAM` is a port to R of the [LiNGAM](https://github.com/cdt15/lingam) package (LiNGAM: Linear Non-Gaussian Acyclic Model), which is available in Python.

This is currently an alpha version under development, and we are releasing it for the purpose of testing and gathering feedback.


## Features

* Implementation of the Direct LiNGAM algorithm
* Stability assessment of causal structures using the bootstrap method
* Visualization of estimation results using DiagrammeR


## Important Notes

* This package does not include all the features of the Python version.
* This package also includes features that are not present in the Python version.


## Installation

You can install the development version of DirectLiNGAM from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("morimotoosamu/DirectLiNGAM")
```


## Requirements

- DiagrammeR
- glmnet

## Usage

### Sample Data

```{r example}
library(DirectLiNGAM)
data(LiNGAM_sample_1000)

m <- matrix(
  c(0.0, 0.0, 0.0, 3.0, 0.0, 0.0,
    3.0, 0.0, 2.0, 0.0, 0.0, 0.0,
    0.0, 0.0, 0.0, 6.0, 0.0, 0.0,
    0.0, 0.0, 0.0, 0.0, 0.0, 0.0,
    8.0, 0.0,-1.0, 0.0, 0.0, 0.0,
    4.0, 0.0, 0.0, 0.0, 0.0, 0.0),
  nrow = 6, byrow = TRUE
  )

colnames(m) <- rownames(m) <- colnames(LiNGAM_sample_1000)

m |>
  plot_adjacency_diagrammer(
  labels      = colnames(LiNGAM_sample_1000),
  title = "True causal structure",
  rankdir     = "TB",
  shape       = "circle"
)
```


### Causal Discovery

独立性の評価はデフォルトでは相互情報量(mutual infomation)を用います。


HSIC(Hilbert-Schmidt Independence Criterion)を使いたい場合は引数で `measure = "kernel"` を指定します。


係数の算出はデフォルトではLASSOを用い、ラムダの選択はAICを用います。


```{r direct_lingam}
model <- direct_lingam(LiNGAM_sample_1000)
```


### Causal Order

推定された因果の順序を確認します。

```{r causal_order}
# index number
model$causal_order

# variable name
colnames(LiNGAM_sample_1000)[model$causal_order]
```

### Estimated Adjacency Matrix

推定された効果の量を確認します。

```{r adjacency_matrix}
B_hat <- model$adjacency_matrix
colnames(B_hat) <- rownames(B_hat) <- colnames(LiNGAM_sample_1000)
round(B_hat, 3)
```


### Plot The Estimated Causal Graph

推定された隣接行列に基づいて、因果グラフを描きます。

余計なpath(x3 to x1, x0 to x2)が推定されていますが、係数はとても小さいです。

Only paths with a coefficient of 0.5 or greater are being drawn.

```{r plot_adjacency}
B_hat |>
  plot_adjacency_diagrammer(
      labels = colnames(LiNGAM_sample_1000),
      title = "Estimated Causal Structure (Direct LiNGAM)",
      rankdir = "TB",
      shape = "ellipse",
      fillcolor = "lightgreen"
      )
```


### Calculating The Total Causal Effect

推定されたすべての総合効果を算出します。

```{r total_effects}
LiNGAM_sample_1000 |>
  estimate_all_total_effects(model) |>
  round(3)
```


### Inference Based On Prior Knowledge

事前知識を用いた実行例です。

#### Specify In The Index
- x3 is an exogenous variable.
- x1, x4, and x5 are sink_variables.

```{r make_prior_knowledge1_index}
pk1 <- make_prior_knowledge(
  n_variables         = 6,
  exogenous_variables = 4,
  sink_variables = c(2, 5, 6)
)

pk1
```

Direct LiNGAM を実行する際に、引数 `prior_knowledge` に事前知識を指定します。

```{r LiNGAM_with_pk_1}
model_pk1 <- LiNGAM_sample_1000 |>
  direct_lingam(prior_knowledge = pk1)

cat("Causal Order: ", colnames(LiNGAM_sample_1000)[model_pk1$causal_order], "\n")
```

結果の隣接行列に基づいて因果グラフを描きます。今度は真の因果構造に非常に近い結果が得られました

```{r LiNGAM_with_pk_2}
B_pk <- model_pk1$adjacency_matrix
colnames(B_pk) <- rownames(B_pk) <- colnames(LiNGAM_sample_1000)
round(B_pk, 3)

plot_adjacency_diagrammer(
  B_pk,
  threshold = 0.5,
  labels      = colnames(LiNGAM_sample_1000),
  title = "Estimated (with Prior Knowledge)",
  rankdir     = "TB",
  shape       = "circle",
  fillcolor   = "lightgreen"
)
```

### Independence between error variables

Calculation of the p-value (default: Spearman)
```{r Independence_between_error_variables1}
result <- LiNGAM_sample_1000 |>
  direct_lingam()

p_vals <- LiNGAM_sample_1000 |>
  get_error_independence_p_values(result)
round(p_vals, 3)
```

### Bootstrap Direct LiNGAM

```{r bootstrap_direct_lingam}
bs_model <- LiNGAM_sample_1000 |>
  bootstrap_lingam(n_sampling = 30L, seed = 42)
```

```{r BootstrapResult}
bs_model
```

ブートストラップの結果

係数

```{r get_causal_direction_counts}
bs_model |>
  get_causal_direction_counts(labels = names(LiNGAM_sample_1000))
```

平均因果効果の隣接行列

```{r get_adjacency_matrix_summary}
bs_adjacency_matrix <- bs_model |>
  get_adjacency_matrix_summary(stat = "median")

bs_adjacency_matrix |>
  round(3)
```

係数の可視化（係数0.5以上のパスを描画）
```{r plot_bootstrap}
bs_adjacency_matrix |>
  plot_adjacency_diagrammer(
    threshold = 0.5,
    labels      = colnames(LiNGAM_sample_1000),
    title = "Estimated (with Bootstrap)",
    rankdir     = "TB",
    shape       = "circle",
    fillcolor   = "lightgreen"
    )
```



ブートストラップ確率の行列

```{r get_probabilities}
bs_model |>
  get_probabilities() 
```

平均総合効果

```{r get_total_causal_effects}
bs_model |>
  get_total_causal_effects()
```

bootstrapの結果を因果グラフに

デフォルトでは50%以上出現しているパスを表示

```{r}
bs_model |>
  plot_bootstrap_probabilities()
```


### sample data 10000

```{r example_10000}
data(LiNGAM_sample_10000)

m <- matrix(c(
  0, 0, 0, 3, 0, 0, 0, 0, 0, 0,
  3, 0, 2, 0, 0, 0, 0, 0, 0, 0,
  0, 0, 0, 6, 0, 0, 0, 0, 0, 0,
  0, 0, 0, 0, 0, 0, 0, 0, 0, 0,
  8, 0, -1, 0, 0, 0, 0, 0, 0, 0,
  4, 0, 0, 0, 0, 0, 0, 0, 0, 0,
  0, 2, 0, 0, 0, 0, 0, 0, 0, 0,
  3, 0, 0, 0, 1.5, 0, 0, 0, 0, 0,
  0, 0, 0.5, 0, 0, 2, 0, 0, 0, 0,
  0, 0, 0, 7, 0, 0, 1, 0, 0, 0
  ),
  nrow = 10, byrow = TRUE
  )

colnames(m) <- rownames(m) <- colnames(LiNGAM_sample_10000)

m |>
  plot_adjacency_diagrammer(
  labels      = colnames(LiNGAM_sample_10000),
  title = "True causal structure",
  rankdir     = "TB",
  shape       = "circle"
)
```



## Licence

MIT License

Original work: Copyright (c) 2019 T.Ikeuchi, G.Haraoka, M.Ide, W.Kurebayashi, S.Shimizu

Portions of this work: Copyright (c) 2026 O.Morimoto


## Feedback

Please submit bug reports and feature requests via GitHub Issues.

