---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```


# lingamr

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html)
<!-- badges: end -->

LiNGAM is a new method for estimating structural equation models or linear Bayesian networks. It is based on using the non-Gaussianity of the data.

This package is a port of the Python lingam package to R.

- [The LiNGAM Project](https://sites.google.com/view/sshimizu06/lingam)
- [lingam](https://github.com/cdt15/lingam)

`lingamr` is a port to R of the [LiNGAM](https://github.com/cdt15/lingam) package (LiNGAM: Linear Non-Gaussian Acyclic Model), which is available in Python.

This is currently an alpha version under development, and we are releasing it for the purpose of testing and gathering feedback.


## Features

* Implementation of the Direct LiNGAM algorithm
* Stability assessment of causal structures using the bootstrap method
* Visualization of estimation results using DiagrammeR


## Important Notes

* This package does not include all the features of the Python version.
* This package also includes features that are not present in the Python version.


## Installation

You can install the development version of DirectLiNGAM from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("morimotoosamu/lingamr")
```


## Requirements

- DiagrammeR
- glmnet

## Usage

```{r library}
library(lingamr)
```

### Sample Data

```{r sample1_matrix6}
x1k <- generate_lingam_sample_6(n = 1000)

x1k$true_adjacency |>
  plot_adjacency(
  labels  = colnames(x1k$data),
  title   = "True causal structure",
  rankdir = "TB",
  shape   = "circle"
)
```


### Causal Discovery

独立性の評価はデフォルトでは相互情報量(mutual infomation)を用います。

パス係数の算出には適応的LASSO回帰が使われます。

```{r lingam_direct}
model <- x1k$data |>
  lingam_direct()
```


### Causal Order

推定された因果の順序を確認します。

```{r causal_order}
# index number
model$causal_order

# variable name
colnames(x1k$data)[model$causal_order]
```

### Estimated Adjacency Matrix

推定された効果の量を確認します。

```{r adjacency_matrix}
model$adjacency_matrix |>
  round(3)
```


### Plot The Estimated Causal Graph

推定された隣接行列に基づいて、因果グラフを描きます。

```{r plot_adjacency}
model$adjacency_matrix |>
  plot_adjacency(
      labels = colnames(model$adjacency_matrix),
      title = "Estimated Causal Structure (Direct LiNGAM)",
      rankdir = "TB",
      shape = "ellipse",
      fillcolor = "lightgreen"
      )
```


### Calculating The Total Causal Effect

推定されたすべての総合効果を算出します。

```{r total_effects}
x1k$data |>
  estimate_all_total_effects(model) |>
  round(3)
```


### Inference Based On Prior Knowledge

事前知識を用いた実行例です。

#### Specify In The Index
- 変数の数は6個
- x3 is an exogenous variable.
- x1, x4, and x5 are sink_variables.
- x0 to x2 are no path.

```{r make_prior_knowledge1_index}
pk1 <- make_prior_knowledge(
  n_variables         = 6,
  exogenous_variables = 4,
  sink_variables = c(2, 5, 6),
  no_paths = list(c(3, 1), c(1, 3))
)

pk1
```

Direct LiNGAM を実行する際に、引数 `prior_knowledge` に事前知識を指定します。

```{r LiNGAM_with_pk_1}
model_pk1 <- x1k$data |>
  lingam_direct(prior_knowledge = pk1, lambda = "BIC")

cat("Causal Order: ", colnames(x1k$data)[model_pk1$causal_order], "\n")
```

結果の隣接行列に基づいて因果グラフを描きます。

```{r LiNGAM_with_pk_2}
model_pk1$adjacency_matrix |>
  round(3)

model_pk1$adjacency_matrix |>
  plot_adjacency(
    labels      = colnames(model_pk1$adjacency_matrix),
    title = "Estimated (with Prior Knowledge)",
    rankdir     = "TB",
    shape       = "circle",
    fillcolor   = "lightgreen"
    )
```

### Independence between error variables

LiNGAMでは残差が独立であることを仮定している。

get_error_independence_p_values関数は残差間の独立性の検定のp値を返す。

Calculation of the p-value (default: Spearman)


```{r Independence_between_error_variables1}
result <- x1k$data |>
  lingam_direct()

p_vals <- x1k$data |>
  get_error_independence_p_values(result)

round(p_vals, 3)
```

### 正規性の検定

残差の正規性の検定を行う

```{r nortest}
# Shapiro-Wilk (default)
x1k$data |>
  test_residual_normality(result)
```

QQプロットでも残差の正規性を確認する

```{r qqplot}
x1k$data |>
  plot_residual_qq(result)
```


### Bootstrap Direct LiNGAM

```{r lingam_direct_bootstrap}
bs_model <- x1k$data |>
  lingam_direct_bootstrap(n_sampling = 30L, seed = 42)

bs_model
```


係数

```{r get_causal_direction_counts}
bs_model |>
  get_causal_direction_counts(labels = names(x1k$data))
```

平均因果効果の隣接行列

```{r get_adjacency_matrix_summary}
bs_adjacency_matrix <- bs_model |>
  get_adjacency_matrix_summary(stat = "median")

bs_adjacency_matrix |>
  round(3)
```

係数の可視化（係数0.5以上のパスを描画）
```{r plot_bootstrap}
bs_adjacency_matrix |>
  plot_adjacency(
    threshold = 0.5,
    labels      = colnames(x1k$data),
    title = "Estimated (with Bootstrap)",
    rankdir     = "TB",
    shape       = "circle",
    fillcolor   = "lightgreen"
    )
```



ブートストラップ確率の行列

```{r get_probabilities}
bs_model |>
  get_probabilities() 
```

平均総合効果

```{r get_total_causal_effects}
bs_model |>
  get_total_causal_effects()
```

bootstrapの結果を因果グラフに

デフォルトでは50%以上出現しているパスを表示

```{r}
bs_model |>
  plot_bootstrap_probabilities()
```


### sample data 10変数

大きめのデータセット。10変数、1万行。

```{r sample1_matrix10}
x10k <- generate_lingam_sample_10(n = 10000)

x10k$true_adjacency |>
  plot_adjacency(
  labels  = colnames(x10k$data),
  title   = "True causal structure",
  rankdir = "TB",
  shape   = "circle"
)
```


## Licence

MIT License

Original work: Copyright (c) 2019 T.Ikeuchi, G.Haraoka, M.Ide, W.Kurebayashi, S.Shimizu

Portions of this work: Copyright (c) 2026 O.Morimoto


## References

### Algorithm
- Shimizu, S. et al. (2011). DirectLiNGAM: A direct method for learning 
  a linear non-Gaussian structural equation model. 
  *Journal of Machine Learning Research*, 12, 1225-1248.
### Original Implementation (Python)
- Ikeuchi, T. et al. (2023). Python package for LiNGAM algorithms.
  *Journal of Machine Learning Research*, 24(14), 1-7.  
  https://github.com/cdt15/lingam
### Books
- 清水昌平(2017)『統計的因果探索』講談社.
- 梅津佑太・西村龍映・上田勇祐(2020)『スパース回帰分析とパターン認識』講談社.
- 鈴木譲(2025)『グラフィカルモデルと因果探索100問 with R』共立出版.
### R Packages Referenced
- G. Kikuchi (2020). rlingam
  https://github.com/gkikuchi/rlingam


## Feedback

Please submit bug reports and feature requests via GitHub Issues.

