---
title: "Causal Discovery with lingamr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Discovery with lingamr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
set.seed(0)
```

`lingamr` estimates **causal structure** — which variable causes which, and
how strongly — from purely observational data, using the LiNGAM family of
algorithms (an R port of the Python
[`lingam`](https://github.com/cdt15/lingam) package by the
[Shimizu lab](https://www.shimizulab.org/)).

This vignette explains the core idea behind LiNGAM and walks through a
minimal end-to-end workflow. Detailed guides for every method live on the
package website; see [Where to Go Next](#where-to-go-next).

```{r setup}
library(lingamr)
```


## The Idea: Why Non-Gaussianity Reveals Causal Direction

Correlation alone cannot distinguish "x causes y" from "y causes x": both
models can produce exactly the same covariance matrix. Classical methods
based on second-order statistics therefore return, at best, an *equivalence
class* of structures.

LiNGAM (Linear Non-Gaussian Acyclic Model; Shimizu et al. 2006) resolves the
direction by adding one assumption: the error terms are **non-Gaussian**. The
model for each variable is

$$x_i = \sum_{j:\ \mathrm{parent\ of\ } i} b_{ij}\, x_j + e_i,$$

with mutually independent, non-Gaussian errors $e_i$, arranged in a DAG.
Under these assumptions the full causal structure is **uniquely
identifiable** from observational data — not just up to an equivalence class.

The intuition fits in one experiment. Take the true model
$y = 1.5x + e$ with a *uniform* (non-Gaussian) error, and regress in both
directions:

```{r nongauss_intuition, fig.width = 7, fig.height = 3.5}
n <- 1000
x <- runif(n, -1, 1)
y <- 1.5 * x + runif(n, -1, 1)

res_causal <- residuals(lm(y ~ x)) # correct direction:  x -> y
res_anti   <- residuals(lm(x ~ y)) # reverse direction:  y -> x

oldpar <- par(mfrow = c(1, 2))
plot(x, res_causal, main = "Correct: residual of y ~ x", cex = 0.4)
plot(y, res_anti,   main = "Reverse: residual of x ~ y", cex = 0.4)
par(oldpar)
```

In the correct direction the residual is **independent** of the regressor
(left panel: a featureless band). In the reverse direction it is not (right
panel: the residual's spread depends on $y$). Direct LiNGAM turns this
asymmetry into an algorithm: the variable whose residuals are most
independent of it is the most upstream ("exogenous") one; peel it off,
regress it out, and repeat.

Had the error been Gaussian, both panels would look identical — this is why
non-Gaussianity is essential, and why the *basic* LiNGAM model rests on four
assumptions:

1. **Linear** relationships,
2. an **acyclic** graph (a DAG),
3. **non-Gaussian**, mutually independent errors,
4. **no latent confounder** (all common causes observed),

plus i.i.d. observations. `lingamr` also ships estimators that relax each of
these — see `vignette("method-selection")`.


## A Minimal Workflow

### Estimate

`generate_lingam_sample_6()` generates data from a known 6-variable LiNGAM
model, so we can compare estimates against the truth. `lingam_direct()` runs
Direct LiNGAM — by default, independence is assessed via mutual information
and coefficients are estimated with adaptive LASSO.

```{r workflow_fit}
x1k <- generate_lingam_sample_6(n = 1000)

model <- lingam_direct(x1k$data)
model
```

The two key components of the result:

```{r workflow_inspect}
# The estimated causal order, upstream first
colnames(x1k$data)[model$causal_order]

# The adjacency matrix: B[i, j] is the direct effect of x_j on x_i
round(model$adjacency_matrix, 3)
```

### Visualize

`plot_adjacency()` draws the causal graph; passing the true structure via
`true_B` color-codes the comparison (green = correct, red = false positive,
orange dashed = missed).

```{r workflow_plot}
model$adjacency_matrix |>
  plot_adjacency(
    labels = colnames(x1k$data),
    true_B = x1k$true_adjacency,
    title  = "Estimated vs. true structure"
  )
```

### Intervene: Total Causal Effects

The total causal effect — how much a variable ultimately moves when another
is changed by one unit, through all paths — is what you need to reason about
interventions (a multiple-regression coefficient answers a different
question; see the
[Direct LiNGAM article](https://morimotoosamu.github.io/lingamr/articles/direct-lingam.html)).

```{r workflow_total}
total_effects <- estimate_all_total_effects(x1k$data, model)
round(total_effects, 2)
```

### Check the Assumptions

`summary_lingam()` bundles the two key diagnostics: residuals should be
mutually **independent** (assumption 4) and **non-Gaussian** (assumption 3 —
so normality being *rejected* is good news here).

```{r workflow_summary}
summary_lingam(x1k$data, model)
```

### Quantify Stability

The bootstrap re-runs the estimation on resampled data and reports how often
each edge recurs — low-probability edges should not be over-interpreted:

```{r workflow_bootstrap}
bs <- lingam_direct_bootstrap(x1k$data, n_sampling = 50L, seed = 42)

round(get_probabilities(bs), 2)
```


## Where to Go Next

Which estimator fits your data — time series, latent confounders, nonlinear
relationships, mixed or missing data — is covered in
`vignette("method-selection")`.

Detailed worked examples live on the
[package website](https://morimotoosamu.github.io/lingamr/):

| Article | Covers |
|---|---|
| [Direct LiNGAM in depth](https://morimotoosamu.github.io/lingamr/articles/direct-lingam.html) | Prior knowledge, regression methods, non-Gaussianity experiments, high-dimensional data, failure modes |
| [Bootstrap and diagnostics](https://morimotoosamu.github.io/lingamr/articles/bootstrap-diagnostics.html) | Stability analysis, assumption checks, SEM fit, broom integration |
| [Time series](https://morimotoosamu.github.io/lingamr/articles/time-series.html) | VAR-LiNGAM, VARMA-LiNGAM |
| [Latent confounders](https://morimotoosamu.github.io/lingamr/articles/latent-confounders.html) | BottomUpParceLiNGAM, RCD |
| [Nonlinear methods](https://morimotoosamu.github.io/lingamr/articles/nonlinear.html) | RESIT, CAM-UV |
| [Special data](https://morimotoosamu.github.io/lingamr/articles/special-data.html) | Mixed data (LiM), multiple groups, missing values |

A Japanese translation of this vignette is available as
`vignette("lingamr-ja")`; the website articles are also available in
Japanese.
