---
title: "Bootstrap Stability and Model Diagnostics"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
set.seed(0)
```

A causal discovery estimate is only as good as the assumptions behind it.
This article covers the tools `lingamr` provides to assess an estimated
structure:

- **Assumption checks**: residual independence and non-Gaussianity
  (`get_error_independence_p_values()`, `test_residual_normality()`,
  `summary_lingam()`).
- **Bootstrap stability**: how reproducible are the estimated edges and the
  causal order (`lingam_direct_bootstrap()` and its query functions)?
- **Model fit**: SEM fit measures for the estimated graph
  (`evaluate_model_fit()`).
- **broom integration**: `tidy()` / `glance()` for downstream analysis.

The examples use Direct LiNGAM on the standard 6-variable sample data; the
same workflow applies to the other estimators (their bootstrap variants are
described in the respective articles).

```{r setup}
library(lingamr)

x1k <- generate_lingam_sample_6(n = 1000)
model <- lingam_direct(x1k$data)
```


## Independence between Error Variables

LiNGAM assumes that the residuals are independent.
`get_error_independence_p_values()` returns the p-values from tests of
independence between the residuals.

```{r Independence_between_error_variables1}
p_vals <- x1k$data |>
  get_error_independence_p_values(model)

round(p_vals, 3)
```

Small p-values (dependence between residuals) suggest a latent confounder or
a mis-specified structure — consider the methods in the
[latent confounders article](latent-confounders.html).


## Testing the Normality of Residuals

We test the normality of the residuals. Because LiNGAM assumes non-Gaussianity,
having normality **rejected** (a small p-value) is consistent with the model's
assumptions.

```{r nortest}
# Shapiro-Wilk (default)
x1k$data |>
  test_residual_normality(model)
```

We also check the normality of the residuals with a QQ plot.

```{r qqplot}
x1k$data |>
  plot_residual_qq(model)
```

If normality is *not* rejected, the causal direction may be unidentifiable —
see the non-Gaussianity experiments in the
[Direct LiNGAM article](direct-lingam.html).


## Model Summary

`summary_lingam()` runs the residual independence test and the normality test
together, letting you review at a glance how well the two assumptions LiNGAM
relies on hold (that the residuals are mutually independent, and that the
residuals are non-Gaussian). Instead of calling
`get_error_independence_p_values()` and `test_residual_normality()` separately,
you can survey the diagnostics in one place.

```{r summary_lingam}
x1k$data |>
  summary_lingam(model)
```


## Bootstrap Direct LiNGAM

We assess the reliability of the model using the bootstrap method.

```{r lingam_direct_bootstrap}
bs_model <- x1k$data |>
  lingam_direct_bootstrap(n_sampling = 100L, seed = 42)

bs_model
```

When the number of iterations or variables is large, specifying
`parallel = TRUE` lets it run faster on multiple cores. The number of cores is
specified with `n_cores` (when unspecified, it is capped at 2 cores for safety).

```{r lingam_direct_bootstrap_parallel, eval = FALSE}
bs_model <- x1k$data |>
  lingam_direct_bootstrap(
    n_sampling = 100L,
    seed       = 42,
    parallel   = TRUE,
    n_cores    = 4L
  )
```

Note that parallel execution uses L'Ecuyer's parallel random number streams, so
results are reproducible given the same `seed` and the same `n_cores`, but they
will not numerically match the results of sequential execution
(`parallel = FALSE`).


### Inspecting the Bootstrap Results

From the bootstrap results, we compute the frequency of occurrence of each path
and the mean of the coefficients.

```{r get_causal_direction_counts}
bs_model |>
  get_causal_direction_counts(labels = names(x1k$data))
```


### Adjacency Matrix of Mean Causal Effects

We construct an adjacency matrix from the bootstrap results.

```{r get_adjacency_matrix_summary}
bs_adjacency_matrix <- bs_model |>
  get_adjacency_matrix_summary(stat = "median")

bs_adjacency_matrix |>
  round(3)
```

We visualize the estimated adjacency matrix.
```{r plot_bootstrap}
bs_adjacency_matrix |>
  plot_adjacency(
    labels    = colnames(x1k$data),
    title     = "Estimated (with Bootstrap)",
    rankdir   = "TB",
    shape     = "circle",
    fillcolor = "lightgreen"
  )
```


### Matrix of Path Occurrence Frequencies

We compute the matrix of occurrence frequencies for each path.

```{r get_probabilities}
bs_model |>
  get_probabilities()
```


### Mean Total Effects

We compute the mean total effect of each path.

```{r get_total_causal_effects}
bs_model |>
  get_total_causal_effects()
```

We turn the bootstrap results into a causal graph. By default, only paths that
occur in at least 50% of samples are shown.

```{r plot_bootstrap_probabilities}
bs_model |>
  plot_bootstrap_probabilities()
```


### Stability of the Causal Order

`get_causal_order_stability()` aggregates the causal orders estimated in each
bootstrap sample and quantifies how stable the order is. It returns the rank
distribution of each variable, the precedence probability for variable pairs
(`P[i, j]` = the fraction of samples in which variable i came upstream of j),
and an overall stability score (0 = random, 1 = identical across all samples).

```{r get_causal_order_stability}
bs_model |>
  get_causal_order_stability(labels = names(x1k$data))
```

> **Caveat:** bootstrap stability does not guarantee correctness. When a model
> assumption is violated, the wrong structure can be reproduced *stably* — see
> the measurement error paradox in the
> [Direct LiNGAM article](direct-lingam.html) for a worked example.


## Evaluating Model Fit

`evaluate_model_fit()` treats an estimated adjacency matrix as a structural
equation model (SEM) and reports standard SEM fit measures (CFI, RMSEA,
AIC/BIC, etc.) via the `lavaan` package (an optional dependency; install it
with `install.packages("lavaan")`). This is useful for judging whether an
estimated causal graph is consistent with the data, independent of how it
was estimated.

```{r evaluate_model_fit, eval = requireNamespace("lavaan", quietly = TRUE)}
sample6 <- generate_lingam_sample_6()
fit_result <- lingam_direct(sample6$data, reg_method = "ols")

# fit measures for the estimated graph
evaluate_model_fit(fit_result, sample6$data)
```

Reversing the direction of every edge produces a mis-specified model, and
its fit measures are visibly worse (lower CFI, higher RMSEA):

```{r evaluate_model_fit_reversed, eval = requireNamespace("lavaan", quietly = TRUE)}
reversed_adjacency <- t(fit_result$adjacency_matrix)
evaluate_model_fit(reversed_adjacency, sample6$data)
```


## Integration with broom (tidy / glance)

Estimation results can be converted to a data.frame with the `broom`-compatible
`tidy()` / `glance()`, making integration with `ggplot2` and `dplyr` easy.
`tidy()` returns an edge list (`from`, `to`, `estimate`), and `glance()` returns
a one-row summary of the whole model. `tidy()` also works on bootstrap results,
in which case it returns the occurrence frequencies for each direction, etc.

```{r broom}
# Convert the estimated adjacency matrix to an edge list
tidy(model)

# One-row summary of the whole model
glance(model)

# Direction-wise summary of the bootstrap results (variable names via labels)
tidy(bs_model, labels = names(x1k$data))
```


## Related Articles

- [Method selection guide](method-selection.html)
- [Direct LiNGAM in depth](direct-lingam.html) — including the measurement
  error paradox, where the bootstrap is stably wrong
- [Time series](time-series.html), [latent confounders](latent-confounders.html),
  [special data](special-data.html) — each method's own bootstrap variant
