---
title: "Direct LiNGAM in Depth"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
set.seed(0)
```

This article is a deep dive into `lingam_direct()`, the core estimator of
`lingamr` for i.i.d. continuous data: how to read its output, estimate total
causal effects, incorporate prior knowledge, choose the regression method,
and understand when and why it fails. For a first tour of the package, start
with `vignette("lingamr")`; to pick the right method for your data, see the
[method selection guide](method-selection.html).

```{r setup}
library(lingamr)
```


## Sample Data

`lingamr` provides five general-purpose sample data generators (further
method-specific generators exist for the other estimators). Each returns a
list containing `data` (a data frame) and `true_adjacency` (the true
adjacency matrix).

| Function | Variables | Default n | Characteristics |
|------|:------:|:-----------:|------|
| `generate_lingam_sample_6()` | 6 | 1,000 | Standard fixed structure. The main example in this article |
| `generate_lingam_sample_10()` | 10 | 1,000 | An extension of the 6-variable case (used in [A Larger Dataset](#a-larger-dataset-10-variables)) |
| `generate_lingam_hard_sample()` | 9 | 200 | A difficult setting with strong multicollinearity (used in [Strong Multicollinearity](#strong-multicollinearity-init_method)) |
| `generate_lingam_large_sample()` | variable | 1,000 | A random sparse DAG with an arbitrary number of variables (used in [The Scalability Wall](#when-there-are-many-variables-the-scalability-wall)) |
| `generate_lingam_paradox_data()` | 4 | 2,000 | The measurement error paradox (used in [The Paradox Example](#a-case-where-directlingam-struggles-the-measurement-error-paradox)) |

### generate_lingam_sample_6()

`generate_lingam_sample_6()` returns artificial data following a 6-variable
LiNGAM model, together with its true adjacency matrix.
The data is stored in `data` and the adjacency matrix in `true_adjacency`.

```{r sample1_matrix6_1}
x1k <- generate_lingam_sample_6(n = 1000)

x1k$data |>
  head()
```

```{r sample1_matrix6_2}
x1k$true_adjacency
```

`plot_adjacency()` draws a causal graph based on the adjacency matrix.

```{r sample1_matrix6_3}
x1k$true_adjacency |>
  plot_adjacency(
    labels  = colnames(x1k$data),
    title   = "True causal structure",
    rankdir = "TB",
    shape   = "circle"
  )
```


## Causal Discovery

`lingam_direct()` runs Direct LiNGAM. By default, independence is assessed
using mutual information, and path coefficients are computed with adaptive
LASSO regression.

```{r lingam_direct}
model <- x1k$data |>
  lingam_direct()
```

To use HSIC for assessing independence, set the `measure` argument to
"kernel". HSIC is computationally expensive; for `n > 1000`,
`lingam_direct()` automatically switches to a low-rank approximation.


### Causal Order

The estimated causal order is stored in `causal_order` as index numbers.
```{r causal_order}
# index number
model$causal_order

# variable name
colnames(x1k$data)[model$causal_order]
```


### Estimated Adjacency Matrix

We inspect the estimated effect magnitudes. By default, the regression
coefficients from adaptive LASSO regression are used.

```{r adjacency_matrix}
model$adjacency_matrix |>
  round(3)
```


### Drawing the Causal Graph

We draw the causal graph based on the adjacency matrix estimated by Direct LiNGAM.

```{r plot_adjacency}
model$adjacency_matrix |>
  plot_adjacency(
    labels    = colnames(model$adjacency_matrix),
    title     = "Estimated Causal Structure (Direct LiNGAM)",
    rankdir   = "TB",
    shape     = "ellipse",
    fillcolor = "lightgreen"
  )
```


### Comparing the Estimated and True Structures

When the true structure is known, as with sample data, you can pass the true
adjacency matrix to the `true_B` argument of `plot_adjacency()` to color-code
the estimated edges by comparing them against the true structure. This lets you
assess estimation accuracy at a glance, which is useful for validating methods
or for educational purposes.

- **Green (solid)**: correctly detected edges (estimated and true)
- **Red (solid)**: falsely detected edges (estimated but not true)
- **Orange (dashed)**: missed edges (true but not estimated; the true coefficient is shown)

```{r plot_adjacency_compare}
model$adjacency_matrix |>
  plot_adjacency(
    labels  = colnames(model$adjacency_matrix),
    true_B  = x1k$true_adjacency,
    title   = "Estimated vs. True Structure",
    rankdir = "TB",
    shape   = "ellipse"
  )
```


### Static Plotting with ggplot2

While `plot_adjacency()` returns an interactive HTML figure via DiagrammeR,
`autoplot()` draws the same causal structure as a static, ggplot2-based figure.
This is stable for image and PDF output in R Markdown / Quarto, and you can
layer ggplot2 functions on top to set themes or titles afterward. Node positions
are computed using `igraph`'s hierarchical layout, so the causal flow generally
runs from top to bottom.

`autoplot()` is a ggplot2 generic, so call it as `ggplot2::autoplot()` or load
it beforehand with `library(ggplot2)` (plotting requires `ggplot2` and `igraph`).

```{r autoplot, eval = requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("igraph", quietly = TRUE)}
ggplot2::autoplot(model)
```


## Total Causal Effect

The **total causal effect** is the overall impact of changing one variable by
one unit, combining the direct path and all indirect paths (paths through
mediating variables).

```{r total_effects}
total_effects <- x1k$data |>
  estimate_all_total_effects(model)

round(total_effects, 3)
```


### Comparison with Multiple Regression Coefficients

Multiple regression coefficients and total causal effects do not agree when
mediating variables are present.

In the true causal structure of `generate_lingam_sample_6()`, there are two
paths from x3 to x1 (there is no **direct** edge from x3 to x1).

- x3 -> x0 -> x1 (indirect effect: 3.0 x 3.0 = **9.0**)
- x3 -> x2 -> x1 (indirect effect: 6.0 x 2.0 = **12.0**)
- **Total causal effect of x3 on x1 = 9.0 + 12.0 = 21.0**

We compare the coefficients from an OLS regression that includes all variables
to predict x1 against the results of `estimate_all_total_effects()`.

```{r total_vs_regression}
# Multiple regression: include all variables to predict x1
lm_coefs <- coef(lm(x1 ~ ., data = x1k$data))

# Comparison (variables causally related to x1: x0, x2, x3)
data.frame(
  variable           = c("x0", "x2", "x3"),
  OLS_coefficient    = round(lm_coefs[c("x0", "x2", "x3")], 3),
  total_causal_effect = round(total_effects["x1", c("x0", "x2", "x3")], 3)
)
```

The OLS coefficient for x3 is nearly **0**. This is because including x0 and x2
(the mediating variables) in the model causes x3's "effect through mediation"
to be absorbed into the coefficients of x0 and x2.

In contrast, the value of x3 from `estimate_all_total_effects()` is **~21**,
which correctly represents how much x1 ultimately changes when x3 is moved by
one unit.

| Question | Metric to use |
|------|------------|
| "How does x1 change if I move x3 while holding x0 and x2 fixed?" | OLS multiple regression coefficient |
| "How does x1 change if I move x3, through all paths?" | Total causal effect |

When you want to know "the ultimate impact of intervening on a variable," use
the total causal effect rather than the multiple regression coefficient.


### Effect Between a Single Pair of Variables

`estimate_all_total_effects()` returns the full matrix of pairwise effects at
once. When only one specific `from -> to` pair is needed,
`estimate_total_effect()` computes it directly; `from_index` and `to_index`
accept either a variable name or a 1-based index.

```{r total_effect_single}
te_x3_x1 <- x1k$data |>
  estimate_total_effect(model, from_index = "x3", to_index = "x1")

round(te_x3_x1, 3)

# Same value as the x3 column of the total_effects matrix computed above
isTRUE(all.equal(te_x3_x1, total_effects["x1", "x3"], check.attributes = FALSE))
```


## Inference with Prior Knowledge

With `make_prior_knowledge()`, you can incorporate domain knowledge about the
causal relationships among variables into Direct LiNGAM. This narrows the search
space and stabilizes estimation.

### Format of the Prior Knowledge Matrix

`make_prior_knowledge()` returns a $p \times p$ integer matrix. It uses the
indexing convention **row = effect variable (to), column = cause variable (from)**,
the same convention as the adjacency matrix.

| Value | Meaning |
|----|------|
| `-1` | Unknown (default; Direct LiNGAM searches freely) |
| `0` | This edge does not exist |
| `1` | This edge definitely exists |

The following shows how each argument affects the matrix.

| Argument | Value set | Meaning |
|------|------------|------|
| `exogenous_variables` | the entire **row** of the specified variable -> `0` | Receives no influence from any variable (root variable) |
| `sink_variables` | the entire **column** of the specified variable -> `0` | Exerts no influence on any variable (sink variable) |
| `paths` | `pk[to, from] = 1` | Specifies that this edge exists |
| `no_paths` | `pk[to, from] = 0` | Specifies that this edge does not exist |

Variables can be specified either by **1-based index** or by **variable name**
(which requires the `labels` argument).

### Usage Example

We supply domain knowledge about the true structure of `generate_lingam_sample_6()`.

- **x3** (index 4) is exogenous -- it receives no influence from any other variable
- **x1, x4, x5** (indices 2, 5, 6) are sink variables -- they exert no influence on other variables
- **Between x0 and x2** there is no path (in either direction)

#### Specifying by Index

```{r make_prior_knowledge1_index}
pk1 <- make_prior_knowledge(
  n_variables         = 6,
  exogenous_variables = 4,          # x3
  sink_variables      = c(2, 5, 6), # x1, x4, x5
  no_paths            = list(c(3, 1), c(1, 3)) # no x2<->x0
)

pk1
```

How to read the matrix: if `pk1["x1", "x3"]` is `-1`, then "x3->x1 is unknown
(LiNGAM searches for it)"; if `0`, then "x3->x1 does not exist".

#### Specifying by Variable Name

Passing `labels` lets you specify by variable name. This improves readability
and is robust to adding or reordering columns.

```{r make_prior_knowledge1_name}
pk1_named <- make_prior_knowledge(
  n_variables         = 6,
  exogenous_variables = "x3",
  sink_variables      = c("x1", "x4", "x5"),
  no_paths            = list(c("x2", "x0"), c("x0", "x2")),
  labels              = colnames(x1k$data)
)

# Equivalent in content to pk1
identical(pk1, pk1_named)
```

### Running Direct LiNGAM with Prior Knowledge

Simply pass it to the `prior_knowledge` argument and it is reflected in the search.

```{r LiNGAM_with_pk_1}
model_pk1 <- x1k$data |>
  lingam_direct(prior_knowledge = pk1, lambda = "BIC")

cat("Causal Order: ", colnames(x1k$data)[model_pk1$causal_order], "\n")
```

```{r LiNGAM_with_pk_2}
model_pk1$adjacency_matrix |>
  round(3)

model_pk1$adjacency_matrix |>
  plot_adjacency(
    labels    = colnames(model_pk1$adjacency_matrix),
    title     = "Estimated (with Prior Knowledge)",
    rankdir   = "TB",
    shape     = "circle",
    fillcolor = "lightgreen"
  )
```


## Choosing a Regression Method (reg_method)

In Direct LiNGAM, the adjacency matrix is estimated by regression after the
causal order is determined. The `reg_method` argument selects that regression method.

| `reg_method` | `glmnet` | Sparsification | Characteristics |
|---|---|---|---|
| `"ols"` | Not required | None | Estimates all edges. For sanity checks or environments without the package |
| `"lasso"` | Required | Yes | Shrinks weak edges to 0 |
| `"adaptive_lasso"` | Required | Yes (strong) | **Default**. Has the oracle property -- reliably sets truly zero edges to 0 |
| `"ridge"` | Required | None | Stabilizes coefficients with $\ell_2$ regularization. Robust to multicollinearity. Does not sparsify |

The oracle property is the theoretical guarantee that "the true structure can be
reliably recovered as the sample size grows," so `"adaptive_lasso"` is usually
recommended.

### Comparison of the Four Methods

```{r reg_method_compare}
fit_ols    <- lingam_direct(x1k$data, reg_method = "ols")
fit_lasso  <- lingam_direct(x1k$data, reg_method = "lasso",          lambda = "BIC")
fit_alasso <- lingam_direct(x1k$data, reg_method = "adaptive_lasso", lambda = "BIC")
fit_ridge  <- lingam_direct(x1k$data, reg_method = "ridge",          lambda = "BIC")

# Compare the adjacency matrices side by side
round(fit_ols$adjacency_matrix,    3)
round(fit_lasso$adjacency_matrix,  3)
round(fit_alasso$adjacency_matrix, 3)
round(fit_ridge$adjacency_matrix,  3)
```

OLS and Ridge tend to leave nonzero coefficients on all edges, whereas LASSO and
Adaptive LASSO shrink superfluous edges to 0. Ridge reduces the **magnitude** of
coefficients but does not set them to zero.

### Choosing lambda (common to LASSO / Adaptive LASSO)

The choice of penalty strength $\lambda$ directly determines the sparsity of the estimate.

| `lambda` | Method | Sparsity | Use |
|---|---|---|---|
| `"BIC"` | Information criterion | Highest | **Default**. Stable even with small samples |
| `"AIC"` | Information criterion | High | Leaves slightly more edges than BIC |
| `"lambda.min"` | CV (minimum prediction error) | Low | Prioritizes predictive accuracy. More edges |
| `"lambda.1se"` | CV (1SE rule) | Medium to high | Robust CV variant |
| `"oracle"` | Analytic formula (adaptive_lasso only) | - | $\lambda = 5 / n^{1.75}$. Guarantees the theoretical oracle property |

```{r reg_method_lambda}
# Compare BIC (default, sparsest) and lambda.min (minimum prediction error)
fit_bic     <- lingam_direct(x1k$data, lambda = "BIC")
fit_lam_min <- lingam_direct(x1k$data, lambda = "lambda.min")

# Number of nonzero edges
sum(fit_bic$adjacency_matrix     != 0)
sum(fit_lam_min$adjacency_matrix != 0)
```


## Strong Multicollinearity (init_method) {#strong-multicollinearity-init_method}

`generate_lingam_hard_sample()` builds five root variables, `x0` through
`x4`, that all share one strong common factor (`collinearity`, default 0.95),
so they are highly correlated with each other even though none of them
causes any other. Four downstream variables, `x5` through `x8`, are linear
combinations of these roots (`x8` also depends on `x5` and `x6`). This is
harder than ordinary multicollinearity among the predictors of a single
response: here the near-collinear variables are themselves exogenous, which
breaks DirectLiNGAM's assumption that the error terms of root variables are
mutually independent.

```{r hard_generate}
hard <- generate_lingam_hard_sample()

head(hard$data)
hard$true_adjacency

# x0-x4 share a common factor and are therefore highly correlated
round(cor(hard$data[c("x0", "x1", "x2", "x3", "x4")]), 3)
```

We run Direct LiNGAM with its defaults.

```{r hard_fit}
fit_hard <- lingam_direct(hard$data)

colnames(hard$data)[fit_hard$causal_order]
```

Because `x0` through `x4` are mutually dependent, the estimated causal order
does not fully respect the true structure. We check this directly: for every
true edge `from -> to`, `from` should precede `to` in the estimated order.

```{r hard_order_check}
true_edges <- which(hard$true_adjacency != 0, arr.ind = TRUE)
order_pos  <- match(seq_len(ncol(hard$data)), fit_hard$causal_order)

edge_check <- data.frame(
  from            = colnames(hard$data)[true_edges[, "col"]],
  to              = colnames(hard$data)[true_edges[, "row"]],
  order_respected = order_pos[true_edges[, "col"]] < order_pos[true_edges[, "row"]]
)

edge_check
sum(edge_check$order_respected)   # number of true edges correctly ordered
```

Only 6 of the 11 true edges are correctly ordered. The rest -- all edges into
`x6` or `x8` -- are missed or estimated in the reverse direction, because
their true parents were placed too late in the estimated order. This is the
failure mode `generate_lingam_hard_sample()` is designed to expose: correlated
exogenous variables can break the causal order search before regression even
starts.

Even so, the `x0 -> x7` edge happens to be ordered correctly, which lets us
isolate the effect of `init_method` on the regression step alone.
`estimate_total_effect()` estimates Adaptive LASSO's initial weights with OLS
by default (`init_method = "ols"`); `init_method = "ridge"` uses
cross-validated Ridge regression instead, which stays stable under
multicollinearity.

```{r hard_init_method}
set.seed(0)
te_hard_ols <- hard$data |>
  estimate_total_effect(fit_hard, from_index = "x0", to_index = "x7", init_method = "ols")

te_hard_ridge <- hard$data |>
  estimate_total_effect(fit_hard, from_index = "x0", to_index = "x7", init_method = "ridge")

round(c(ols = unname(te_hard_ols), ridge = unname(te_hard_ridge)), 3)
```

The true coefficient of `x0 -> x7` is **2.0**. The OLS-initialized estimate
overshoots substantially, while the Ridge-initialized estimate lands much
closer to the truth -- the improvement `generate_lingam_hard_sample()`'s
documentation attributes to Ridge-initialized Adaptive LASSO under strong
multicollinearity. Keep in mind that `init_method` only affects this
regression stage: it does not change the causal order search, so it cannot
fix the ordering problem shown above.


## The Non-Gaussianity Assumption

The theoretical heart of LiNGAM is the assumption that **the error terms follow
a non-Gaussian distribution**. When the errors are Gaussian, the **direction** of
causation becomes fundamentally unidentifiable (a reverse-direction model that
explains the same distribution exists), and the estimates are unreliable.

We verify this difference in practice by switching the error distribution with
the `noise_dist` argument of `generate_lingam_sample_6()`. The true structure is
as follows (the root is x3).

```{r nongauss_true}
set.seed(0)
truth <- generate_lingam_sample_6(noise_dist = "uniform")

truth$true_adjacency |>
  round(1)
```

The causal graph of the true structure:

```{r nongauss_true_plot}
truth$true_adjacency |>
  plot_adjacency(
    labels = colnames(truth$data),
    title  = "True structure"
  )
```

### Non-Gaussian Errors (Uniform Distribution) -- When It Works

```{r nongauss_uniform}
fit_uniform <- lingam_direct(truth$data)

# Estimated causal order (the true root x3 comes first)
colnames(truth$data)[fit_uniform$causal_order]

# The estimated adjacency matrix recovers the true structure almost perfectly
fit_uniform$adjacency_matrix |>
  round(1)
```

The estimated graph matches the true structure. Edges are color-coded against
the truth: green = correct, red = false positive, orange dashed = missed.

```{r nongauss_uniform_plot}
fit_uniform$adjacency_matrix |>
  plot_adjacency(
    labels = colnames(truth$data),
    true_B = truth$true_adjacency,
    title  = "Estimated (uniform errors)"
  )
```

### Gaussian Errors -- When It Fails

With the same causal structure, the results break down when the errors are Gaussian.

```{r nongauss_gaussian}
gauss <- generate_lingam_sample_6(noise_dist = "gaussian")
fit_gauss <- lingam_direct(gauss$data)

# The causal order does not match the true structure (root x3 does not come first)
colnames(gauss$data)[fit_gauss$causal_order]

fit_gauss$adjacency_matrix |>
  round(1)
```

Compared with the true structure, many edges are wrong (red) or missed (orange
dashed) -- the same color coding as above:

```{r nongauss_gaussian_plot}
fit_gauss$adjacency_matrix |>
  plot_adjacency(
    labels = colnames(gauss$data),
    true_B = truth$true_adjacency,
    title  = "Estimated (Gaussian errors)"
  )
```

With non-Gaussian errors the true adjacency matrix is recovered as-is, whereas
with Gaussian errors both the causal order and the coefficients deviate greatly
from the true structure. This is why it is said that "LiNGAM exploits the
non-Gaussianity of the data to determine the direction of causation." When
applying it to real data, it is important to **test the normality of the
residuals** to check whether this assumption holds — see the
[bootstrap and diagnostics article](bootstrap-diagnostics.html).


## A Larger Dataset (10 Variables)

An example of a larger dataset with 10 variables and 10,000 rows.

```{r sample1_matrix10}
x10k <- generate_lingam_sample_10(n = 10000)

x10k$true_adjacency |>
  plot_adjacency(
    labels  = colnames(x10k$data),
    title   = "True causal structure",
    rankdir = "TB",
    shape   = "circle"
  )
```



## Comparing ICA-LiNGAM and Direct LiNGAM

`pcalg::lingam()` is the original LiNGAM algorithm, which estimates the mixing
matrix with FastICA and obtains the causal order and coefficients
(Shimizu et al. 2006). It solves the same problem while taking an approach
independent of `lingam_direct()`.

### Running Both Algorithms

We analyze the same 6-variable dataset ($n = 1000$) with both methods.

```{r ica_run, eval = requireNamespace("pcalg", quietly = TRUE)}
d_cmp <- generate_lingam_sample_6(n = 1000, seed = 42)

t_cmp_direct <- system.time(res_cmp_direct <- lingam_direct(d_cmp$data))
t_cmp_ica    <- system.time(res_cmp_ica    <- pcalg::lingam(as.matrix(d_cmp$data)))

cat(sprintf("Direct LiNGAM : %.2f sec\nICA-LiNGAM    : %.2f sec\n",
            t_cmp_direct["elapsed"], t_cmp_ica["elapsed"]))
```

### Comparing the Estimated Coefficients

`$Bpruned` uses the same convention as the lingamr adjacency matrix
(`B[i, j]` = coefficient of $x_j \to x_i$).

```{r ica_compare, eval = requireNamespace("pcalg", quietly = TRUE)}
B_ica <- res_cmp_ica$Bpruned
rownames(B_ica) <- colnames(B_ica) <- names(d_cmp$data)

idx_ica  <- which(abs(B_ica) > 0, arr.ind = TRUE)
tidy_ica <- data.frame(
  from  = colnames(B_ica)[idx_ica[, 2]],
  to    = rownames(B_ica)[idx_ica[, 1]],
  ica   = round(B_ica[idx_ica], 3)
)

tidy_dir <- tidy(res_cmp_direct)
tidy_dir <- data.frame(from = tidy_dir$from, to = tidy_dir$to,
                       direct = round(tidy_dir$estimate, 3))

merge(tidy_dir, tidy_ica, by = c("from", "to"), sort = TRUE)
```

### Comparing the DAG Structures

We compare the structures with a full outer join over all edges and check
consistency with the true DAG.

```{r ica_structure, eval = requireNamespace("pcalg", quietly = TRUE)}
B_true   <- d_cmp$true_adjacency
idx_true <- which(abs(B_true) > 0, arr.ind = TRUE)
true_key <- paste(colnames(B_true)[idx_true[, 2]],
                  rownames(B_true)[idx_true[, 1]], sep = "->")

cmp <- merge(tidy_dir, tidy_ica, by = c("from", "to"), all = TRUE, sort = TRUE)
cmp$truth <- paste(cmp$from, cmp$to, sep = "->") %in% true_key
cmp
```

When the `direct` or `ica` column is `NA`, it means that method did not detect
that edge. `truth = TRUE` indicates an edge that exists in the true DAG.

---

## When There Are Many Variables: The Scalability Wall

At each step, Direct LiNGAM performs independence tests on all remaining pairs
of variables. Since the number of steps is $p$ and the number of tests per step
is at most $p(p-1)$, the total number of independence tests is approximately

$$\sum_{k=1}^{p} k(k-1) \approx \frac{p^3}{3}$$

giving a computational cost of **$O(p^3)$**.
By contrast, the FastICA used by ICA-LiNGAM is $O(p^2 n)$ (with BLAS
optimization), so the gap widens as the number of variables grows.

`generate_lingam_large_sample()` generates random sparse DAG data with a freely
configurable number of variables `p`. Each variable $x_i$ ($i \ge 1$) randomly
has at most `max_parents` parents chosen from $x_0, \ldots, x_{i-1}$. Since the
causal order is guaranteed to follow the index order, the adjacency matrix is
always a **lower triangular matrix**.

### Generating the Data

```{r large_gen}
d20 <- generate_lingam_large_sample(p = 20, n = 1000, seed = 42)

dim(d20$data)                    # 1000 rows x 20 columns
sum(d20$true_adjacency != 0)     # number of true edges (sparse DAG)
d20$true_causal_order            # 0, 1, ..., 19
```

### Comparing Execution Times

When $p$ grows by a factor of 1.5 (10 -> 15), the number of independence tests
grows by a factor of $15^3 / 10^3 \approx 3.4$.

```{r large_timing}
d10 <- generate_lingam_large_sample(p = 10, n = 500, seed = 42)
d15 <- generate_lingam_large_sample(p = 15, n = 500, seed = 42)

t10 <- system.time({ r10 <- lingam_direct(d10$data) })
t15 <- system.time({ r15 <- lingam_direct(d15$data) })

cat(sprintf(
  "p = 10 : %.2f sec\np = 15 : %.2f sec\ntheoretical factor %.1fx vs. observed %.1fx\n",
  t10["elapsed"],
  t15["elapsed"],
  15^3 / 10^3,
  t15["elapsed"] / max(t10["elapsed"], 0.01)
))
```

We run ICA-LiNGAM on the same data to compare speed directly.

```{r large_timing_ica, eval = requireNamespace("pcalg", quietly = TRUE)}
t10_ica <- system.time({ pcalg::lingam(as.matrix(d10$data)) })
t15_ica <- system.time({ pcalg::lingam(as.matrix(d15$data)) })

cat(sprintf(
  "              p = 10   p = 15\nDirect LiNGAM : %5.2f sec  %5.2f sec\nICA-LiNGAM    : %5.2f sec  %5.2f sec\n",
  t10["elapsed"], t15["elapsed"],
  t10_ica["elapsed"], t15_ica["elapsed"]
))
```

The larger $p$ becomes, the more Direct LiNGAM's $O(p^3)$ cost dominates, and the
gap between the two widens. In large-scale settings such as $p = 30$ or $p = 50$,
this trend becomes even more pronounced.

### Checking Estimation Accuracy (p = 10)

Even with a sparse DAG, as long as there are **non-Gaussian errors** (default:
uniform distribution), Direct LiNGAM can recover the correct causal order.

```{r large_causal_order}
# Estimated causal order
r10$causal_order

# Whether it matches the true causal order 0, 1, ..., 9 exactly
all(r10$causal_order == d10$true_causal_order)
```

We convert to an edge list with `tidy()` and inspect the estimated coefficients.

```{r large_tidy}
tidy(r10) |>
  head(10)
```


## High-Dimensional Direct LiNGAM

The $O(p^3)$ independence-test cost shown above becomes a real bottleneck once
$p$ grows into the tens or hundreds, and breaks down entirely once $p > n$
(more variables than observations), where the usual regression-based
adjacency estimation is no longer well defined.

`lingam_high_dim()` implements HighDimDirectLiNGAM (Wang & Drton 2020), a
variant designed for this regime. Instead of pairwise independence tests, it
searches the causal order using moment statistics of non-Gaussianity, computed
from a cached Gram matrix. The algorithm is deterministic (no random
restarts), and it returns the same `LingamResult` object as `lingam_direct()`,
so `print()`, `tidy()`, `plot_adjacency()` and `estimate_total_effect()` all
work unchanged.

```{r high_dim_basic}
hd_sample <- generate_lingam_sample_6(n = 500, seed = 1)
hd_result <- lingam_high_dim(hd_sample$data)

hd_result$causal_order
round(hd_result$adjacency_matrix, 3)
```

When `n_samples <= n_features`, the usual BIC-based Adaptive LASSO cannot be
used to estimate the adjacency matrix, so `lingam_high_dim()` falls back to a
cross-validated LASSO (`glmnet::cv.glmnet`) and emits a warning:

```{r high_dim_wide}
wide_sample <- generate_lingam_large_sample(p = 30, n = 25, seed = 1)
wide_result <- lingam_high_dim(wide_sample$data)

wide_result$causal_order
```

## A Case Where DirectLiNGAM Struggles: The Measurement Error Paradox

Causal discovery methods have assumptions, and when those are violated they may
fail to recover the correct structure. `generate_lingam_paradox_data()` is a
dataset designed to deliberately create such a difficult case. As with the other
sample generators, it returns a list containing `data` and `true_adjacency`.

The true structure of this data is a simple serial chain
**x0 -> x1 -> x2 -> x3** (each coefficient 0.8). However, it has two notable features.

- **Heavy measurement error is added to the root variable x0.** This disrupts the
  independence assessment performed in DirectLiNGAM's first step, causing it to
  choose the root incorrectly and making error propagation more likely.
- All variables are **standardized** with `scale()` (no differences in scale).

```{r paradox_generate}
paradox <- generate_lingam_paradox_data(n = 2000L, seed = 42)

head(paradox$data)

# All variables are standardized (sd = 1)
sapply(paradox$data, sd)
```

We visualize the true causal graph. The coefficient 0.8 is the structural
coefficient on the latent scale before standardization.

```{r paradox_true}
paradox$true_adjacency |>
  plot_adjacency(
    labels  = colnames(paradox$true_adjacency),
    title   = "True causal chain (x0 -> x1 -> x2 -> x3)",
    rankdir = "LR",
    shape   = "circle"
  )
```

Now let us apply Direct LiNGAM.

```{r paradox_fit}
model_p <- lingam_direct(paradox$data)

# Estimated causal order
colnames(paradox$data)[model_p$causal_order]
```

Note that the **head of the estimated causal order is x1, not the true root x0**.
Because of the measurement error on the root, DirectLiNGAM fails to select x0 as
the first exogenous variable.

```{r paradox_adjacency}
model_p$adjacency_matrix |>
  round(3)

model_p$adjacency_matrix |>
  plot_adjacency(
    labels    = colnames(model_p$adjacency_matrix),
    title     = "Estimated structure (paradox data)",
    rankdir   = "LR",
    shape     = "circle",
    fillcolor = "lightpink"
  )
```

While the downstream **x1 -> x2 -> x3** is recovered correctly, the **direction
between x0 and x1 is reversed** (the truth is x0 -> x1, but the estimate is
x1 -> x0), and x0 ends up being treated almost like a sink.

We use the bootstrap to check whether this error occurred by chance or is systematic.

```{r paradox_bootstrap}
bs_paradox <- paradox$data |>
  lingam_direct_bootstrap(n_sampling = 100L, seed = 42)

# Occurrence probability of each direction (row = to, column = from)
bs_paradox |>
  get_probabilities() |>
  round(2)
```

The important point is that the incorrect direction **x1 -> x0** is reproduced
with nearly 100% probability. In other words, this error is not coincidental but
**systematic**, and it appears stably across bootstrap samples.

> **Lesson:** Bootstrap stability (high reproduction probability) does not
> guarantee the *correctness* of the estimate. When the model's assumptions (here,
> the assumption that "upstream variables have no measurement error") are violated,
> the method may **stably** recover an incorrect structure. It is important to
> evaluate results critically, together with tests of residual independence and
> normality and domain knowledge about the data-generating process.


## Related Articles

- [Method selection guide](method-selection.html) — which method fits your data
- [Bootstrap and diagnostics](bootstrap-diagnostics.html) — assessing
  reliability and checking assumptions
- [Time series (VAR-LiNGAM and VARMA-LiNGAM)](time-series.html)
- [Latent confounders (ParceLiNGAM and RCD)](latent-confounders.html)
- [Nonlinear methods (RESIT and CAM-UV)](nonlinear.html)
- [Special data (LiM, MultiGroup, missing values)](special-data.html)
