---
title: "Latent Confounders: ParceLiNGAM and RCD"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
set.seed(0)
```

Standard Direct LiNGAM assumes there is no latent (unobserved) confounder:
any variable that causes two or more of the observed variables must itself be
observed. When that assumption fails, `lingam_direct()` will still return a
full causal order, but silently — some part of it may be wrong, with no
indication of which part.

This article covers the two **linear** methods that address latent
confounders:

- **BottomUpParceLiNGAM** (`lingam_parce()`): returns the variables it could
  not order as a single *unresolved block*, signaling where a latent
  confounder is likely.
- **RCD** (`lingam_rcd()`): estimates each variable's ancestor set and flags
  *specific pairs* that share a latent confounder.

If the causal relationships may also be **nonlinear**, see CAM-UV in the
[nonlinear methods article](nonlinear.html), which handles unobserved
variables in nonlinear models.

```{r setup}
library(lingamr)
```


## Bottom-Up ParceLiNGAM

`lingam_parce()` (BottomUpParceLiNGAM, Tashiro et al. 2014) searches for the
causal order from the sink (most downstream) side, testing at each step
whether a candidate variable's residual is independent of the others. As soon
as that test is rejected, the search stops, and every variable it could not
yet place is returned together as a single **unresolved block** — a signal
that those variables likely share a latent confounder, rather than a
(possibly wrong) guess at their order.

`generate_parce_sample()` generates a 7-variable model in which `x6` is an
unobserved common cause of `x2` and `x3`; only `x0`-`x5` are returned as data.

```{r parce_sample}
# HSIC is O(n^2), so a moderate n keeps this article fast to build
confounded <- generate_parce_sample(n = 500, seed = 1)
head(confounded$data)
confounded$confounded_pair
```

```{r parce_fit}
parce_result <- lingam_parce(confounded$data, reg_method = "ols")
print(parce_result)
```

The causal order's first element is the unresolved block, shown in
parentheses; here it correctly contains `x2` and `x3`. The corresponding
entries of the adjacency matrix are `NA`, while edges among the remaining,
fully-resolved variables are estimated as usual:

```{r parce_block}
parce_result$causal_order[[1]]
parce_result$adjacency_matrix[confounded$confounded_pair, confounded$confounded_pair]
```

Because a confounded variable's true parents cannot be identified,
`estimate_total_effect_parce()` warns and returns `NA` when asked for a total
effect *from* a variable in the unresolved block, but still computes normal
estimates for well-identified pairs:

```{r parce_total_effect}
# from a confounded variable: warns and returns NA
estimate_total_effect_parce(confounded$data, parce_result,
  from_index = confounded$confounded_pair[1], to_index = "x1"
)

# a well-identified pair: a normal numeric estimate
estimate_total_effect_parce(confounded$data, parce_result,
  from_index = "x0", to_index = "x5"
)
```

`lingam_parce_bootstrap()` provides bootstrap stability estimates in the same
style as `lingam_direct_bootstrap()`. `NA` (unresolved) edges are treated as
absent when aggregating, so `get_probabilities()` and the other
`BootstrapResult` query functions work as usual; `get_causal_order_stability()`
is the one exception, since ParceLiNGAM's blocked causal order does not fit
its fixed-length format.

```{r parce_bootstrap}
parce_bs <- lingam_parce_bootstrap(confounded$data,
  n_sampling = 10L, reg_method = "ols", seed = 1, verbose = FALSE
)
get_probabilities(parce_bs)
```


## RCD (Repetitive Causal Discovery)

`lingam_rcd()` (Repetitive Causal Discovery; Maeda and Shimizu 2020) tackles
the same latent-confounder problem as `lingam_parce()`, but from a different
angle: rather than searching for a causal order and giving up on an
**unresolved block** once a test is rejected, RCD directly estimates each
variable's **ancestor set** and then checks individual, parent-free pairs for
a shared latent confounder. This makes RCD's output pair-level (which specific
pairs are confounded) rather than block-level (which set of variables could
not be ordered).

`generate_rcd_sample()` generates a 7-variable model in which `x6` is an
unobserved common cause of `x2` and `x4`; only `x0`-`x5` are returned as data.

```{r rcd_sample}
# HSIC is O(n^2), so a moderate n keeps this article fast to build
rcd_confounded <- generate_rcd_sample(n = 300, seed = 1)
head(rcd_confounded$data)
rcd_confounded$confounded_pair
```

```{r rcd_fit}
rcd_result <- lingam_rcd(rcd_confounded$data)
print(rcd_result)
```

`ancestors_list` gives each variable's estimated ancestors (not a causal
order), and the confounded pair's adjacency-matrix entries are `NA`:

```{r rcd_ancestors}
rcd_result$ancestors_list
rcd_result$adjacency_matrix[rcd_confounded$confounded_pair, rcd_confounded$confounded_pair]
```

As with ParceLiNGAM, `estimate_total_effect_rcd()` warns and returns `NA`
when asked for a total effect *from* a confounded variable:

```{r rcd_total_effect}
# from a confounded variable: warns and returns NA
estimate_total_effect_rcd(rcd_confounded$data, rcd_result,
  from_index = rcd_confounded$confounded_pair[1], to_index = rcd_confounded$confounded_pair[2]
)

# a well-identified pair: a normal numeric estimate
estimate_total_effect_rcd(rcd_confounded$data, rcd_result,
  from_index = "x5", to_index = "x0"
)
```

`lingam_rcd_bootstrap()` provides bootstrap stability estimates; the usual
`BootstrapResult` query functions apply.


## ParceLiNGAM or RCD?

| Aspect | ParceLiNGAM | RCD |
|---|---|---|
| Output granularity | Block-level: a set of variables that could not be ordered | Pair-level: which specific pairs are confounded |
| Main result | Blocked causal order + adjacency matrix with `NA` block | Ancestor sets + adjacency matrix with `NA` pairs |
| Question it answers | "Which part of the ordering can I trust?" | "Exactly which pairs share a hidden common cause?" |

Both rely on HSIC independence tests, whose cost grows as $O(n^2)$ — with
large $n$, consider subsampling. If relationships are nonlinear, use
[CAM-UV](nonlinear.html) instead.


## Related Articles

- [Method selection guide](method-selection.html) — which method fits your data
- [Nonlinear methods (RESIT and CAM-UV)](nonlinear.html) — CAM-UV handles
  latent confounders in nonlinear models
- [Bootstrap and diagnostics](bootstrap-diagnostics.html)
