---
title: "Nonlinear Methods: RESIT and CAM-UV"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  eval = requireNamespace("mgcv", quietly = TRUE)
)
set.seed(0)
```

LiNGAM assumes that causal relationships are **linear**. When a cause acts on
its effect through a nonlinear function (saturation, thresholds, quadratic
dose-response, ...), linear methods can pick the wrong direction while giving
no warning. This article covers the two nonlinear methods in `lingamr`:

- **RESIT** (`lingam_resit()`): nonlinear additive noise models,
  assuming *no* latent confounder.
- **CAM-UV** (`lingam_camuv()`): causal additive models that additionally
  allow **unobserved variables**.

Both methods return 0/1 edge indicators rather than coefficients — with
nonlinear causal functions, a single number cannot summarize "the effect of
$x_j$ on $x_i$", so total-effect estimation is intentionally unavailable.

Both also require the suggested package `mgcv` (for GAM regressions).

```{r setup}
library(lingamr)
```


## When Linear LiNGAM Fails on Nonlinear Data

`generate_resit_sample()` generates a 4-variable chain whose causal functions
are all nonlinear ($x_1 = 3x_0^2 + e$, $x_2 = 2\tanh(x_1) + 0.8x_0^3 + e$,
$x_3 = 1.5\sin(x_2) + e$):

```{r resit_sample}
nonlin <- generate_resit_sample(n = 300, seed = 1)
head(nonlin$data)

# True 0/1 adjacency matrix (row = to, column = from)
nonlin$adjacency_matrix
```

Applying **linear** Direct LiNGAM to this data gives a poor result — the
estimated order need not match the true order `x0, x1, x2, x3`:

```{r linear_fails}
linear_fit <- lingam_direct(nonlin$data)
colnames(nonlin$data)[linear_fit$causal_order]
```

This is not a defect of Direct LiNGAM: its linearity assumption simply does
not hold here. A nonlinear method is required.


## RESIT: Nonlinear Additive Noise Models

**RESIT** (Regression with Subsequent Independence Test; Peters et al. 2014)
assumes a nonlinear additive noise model

$$x_i = f_i(\mathrm{parents}(x_i)) + e_i$$

and recovers the structure in two phases:

1. **Order search**: repeatedly detach the most sink-like variable — the one
   whose regression residual on all remaining variables is least dependent on
   them (smallest HSIC statistic).
2. **Pruning**: for each candidate parent, test whether the residual
   regressed on the other parents is already independent of the parent set;
   if so (HSIC p-value above `alpha`), drop the parent.

```{r resit_fit}
resit_result <- lingam_resit(nonlin$data)
print(resit_result)
```

The estimated causal order and the 0/1 adjacency matrix recover the true
nonlinear structure:

```{r resit_inspect}
colnames(nonlin$data)[resit_result$causal_order]
resit_result$adjacency_matrix
```

Since the entries are edge indicators, `plot_adjacency()` draws an unweighted
graph:

```{r resit_plot}
resit_result$adjacency_matrix |>
  plot_adjacency(
    labels = colnames(nonlin$data),
    title  = "RESIT estimate (edges are 0/1, not coefficients)"
  )
```

### Choosing `alpha`

Note the direction of `alpha` (default 0.01): a parent is removed when the
HSIC p-value *exceeds* `alpha`, so **larger** values make the test stricter
about declaring independence and therefore **keep more edges**.

### Custom Regressors

All internal regressions use a smoothing-spline GAM (`mgcv`) by default. Any
nonlinear regressor can be substituted by passing a function
`function(X, y)` that returns fitted values — for example, random forests:

```{r resit_regressor, eval = FALSE}
rf_regressor <- function(X, y) {
  fit <- randomForest::randomForest(X, y)
  as.numeric(predict(fit, X))
}
lingam_resit(nonlin$data, regressor = rf_regressor)
```

### Bootstrap

`lingam_resit_bootstrap()` works like the other bootstrap functions; the
aggregated "probabilities" are detection frequencies of each 0/1 edge.
RESIT performs $O(p^2)$ GAM fits and HSIC tests ($O(n^2)$ each) *per
resample*, so keep `n_sampling` moderate:

```{r resit_bootstrap}
resit_bs <- lingam_resit_bootstrap(nonlin$data,
  n_sampling = 20L, seed = 1, verbose = FALSE
)
get_probabilities(resit_bs)
```


## CAM-UV: Nonlinear Models with Unobserved Variables

RESIT assumes all relevant variables are observed. **CAM-UV**
(Maeda and Shimizu 2021) drops that assumption: it identifies each variable's
direct parents by scanning variable subsets and testing residual
independence, and reports variable pairs whose dependence cannot be explained
by any identified edge as connected through an unobserved variable:

- **UCP** (unobserved causal path): a directed path through an unobserved
  intermediate variable.
- **UBP** (unobserved backdoor path): a common unobserved ancestor
  (a latent confounder).

Such pairs are marked `NA` in the adjacency matrix rather than being forced
into a possibly wrong orientation.

`generate_camuv_sample()` follows the Python tutorial's structure: among six
observed variables, `u1` (latent) confounds `x3` and `x4` (UBP), and `u2`
(latent) mediates the path from `x2` to `x5` (UCP):

```{r camuv_sample}
camuv_dat <- generate_camuv_sample(n = 500, seed = 1)
head(camuv_dat$data)

# The UBP (x3-x4) and UCP (x2-x5) entries of the true matrix are NA
camuv_dat$adjacency_matrix
camuv_dat$confounded_pairs
```

```{r camuv_fit}
camuv_result <- lingam_camuv(camuv_dat$data)
print(camuv_result)
```

Like `lingam_rcd()`, there is no causal order; the result holds each
variable's direct parents and the flagged pairs:

```{r camuv_inspect}
camuv_result$parents_list
camuv_result$confounded_pairs
camuv_result$adjacency_matrix
```

### Tuning

- `num_explanatory_vals` (default 2) bounds the size of the variable subsets
  scanned for parents. Larger values increase statistical power but the cost
  grows combinatorially.
- `independence = "fcorr"` switches the independence measure from the HSIC
  test to the F-correlation with threshold `ind_corr`; it is only supported
  with `num_explanatory_vals = 2`.
- Prior knowledge is given as pairs `c(i, j)` meaning "variable i cannot
  cause variable j" (1-based indices).

There is no bootstrap variant, matching the Python implementation.


## RESIT or CAM-UV?

| Aspect | RESIT | CAM-UV |
|---|---|---|
| Latent confounders | Not allowed | Detected and reported (UCP/UBP pairs) |
| Output | Causal order + 0/1 adjacency matrix | Parents list + adjacency matrix with `NA` pairs |
| Bootstrap | `lingam_resit_bootstrap()` | Not available |
| Cost | $O(p^2)$ GAM fits + HSIC tests | Combinatorial in `num_explanatory_vals` |

Start with RESIT when you can argue that all major common causes are
observed; use CAM-UV when latent confounding is plausible. Both build
$n \times n$ Gram matrices per HSIC test, so neither is recommended for $n$
in the thousands — subsample first if needed.

If the relationships are linear and you suspect latent confounders, the
linear methods in the
[latent confounders article](latent-confounders.html) (ParceLiNGAM, RCD) are
cheaper and return coefficient estimates.


## Related Articles

- [Method selection guide](method-selection.html) — which method fits your data
- [Latent confounders (ParceLiNGAM and RCD)](latent-confounders.html) —
  linear counterparts
- [Bootstrap and diagnostics](bootstrap-diagnostics.html)
