---
title: "Special Data: Mixed Variables, Multiple Groups, and Missing Values"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
set.seed(0)
```

Direct LiNGAM assumes a single, complete data matrix of continuous variables.
This article covers three extensions for data that break those assumptions:

- **LiM** (`lingam_lim()`): data mixing continuous and discrete variables
  (binary or Poisson counts).
- **MultiGroup Direct LiNGAM** (`lingam_multi_group()`): several datasets that
  share a causal structure but not coefficient values.
- **Missing data** (`bootstrap_with_imputation()`): bootstrap combined with
  multiple imputation for data containing `NA`.

```{r setup}
library(lingamr)
```


## LiNGAM for Mixed Data (LiM)

Direct LiNGAM assumes every variable is continuous. `lingam_lim()` relaxes
this assumption and estimates a causal structure from data containing a
mixture of continuous and discrete variables, following
Zeng et al. (2022). It combines a NOTEARS-style continuous optimization
(the "global" phase) with a combinatorial local search over edge
directions, pruning, and edge addition (the "local" phase).

`generate_lim_sample()` generates a small dataset with a known causal chain
of continuous and discrete variables: `x1` (continuous) -> `x2` (discrete)
-> `x3` (continuous).

```{r lim_sample}
set.seed(1)
lim_dat <- generate_lim_sample(n = 2000)
head(lim_dat$data)
lim_dat$is_continuous
```

`lingam_lim()` requires `is_continuous`, a logical vector marking which
columns are continuous (`TRUE`) versus discrete (`FALSE`).
Because the optimization starts from a random initial point,
reproducibility requires `set.seed()`.

```{r lim_fit}
lim_result <- lingam_lim(lim_dat$data, is_continuous = lim_dat$is_continuous)
print(lim_result)
```

As with `lingam_direct()`, `adjacency_matrix` follows the
`B[i, j]` = j -> i convention (row = to, column = from), and `causal_order`
lists the estimated topological order as 1-based indices.

```{r lim_order}
colnames(lim_dat$data)[lim_result$causal_order]
```

### Count Variables (Poisson)

Discrete variables are binary (0/1) by default; setting `is_poisson = TRUE`
treats them as Poisson-distributed counts instead (the local search phase
then scores them with Poisson regression log-likelihoods).
`generate_lim_sample(is_poisson = TRUE)` generates a matching count-data
example:

```{r lim_poisson}
set.seed(2)
lim_pois <- generate_lim_sample(n = 2000, is_poisson = TRUE)
head(lim_pois$data)

lim_pois_result <- lingam_lim(lim_pois$data,
  is_continuous = lim_pois$is_continuous, is_poisson = TRUE
)
print(lim_pois_result)
```

See `?lingam_lim` for details on the local phase's edge-weight convention
and its numeric differences from the Python implementation.


## Multi-Group Direct LiNGAM

`lingam_direct()` fits a single dataset. When data comes from several
sources that plausibly share the same causal structure but not the same
strength of effect (e.g. the same study run at multiple sites, or the same
process observed in different time periods), `lingam_multi_group()` jointly
estimates a **common causal order** across all groups while still allowing
each group its own adjacency matrix (structural coefficients), following
Shimizu (2012).

`generate_multi_group_sample()` generates two datasets that share the causal
structure of `generate_lingam_sample_6()` but with slightly different
coefficients per group.

```{r multi_group_sample}
mg <- generate_multi_group_sample(n = c(1000, 1000), seed = 42)
lapply(mg$data_list, head, 3)
```

```{r multi_group_fit}
mg_result <- lingam_multi_group(mg$data_list, reg_method = "ols")
print(mg_result)
```

`causal_order` is shared by all groups; `adjacency_matrices` holds one
matrix per group, each following the usual `B[i, j]` = j -> i convention.

To analyze a single group with the rest of `lingamr`'s single-group
tooling (total causal effects, independence tests, plotting), extract it
as a plain `LingamResult` with `get_group_result()`:

```{r multi_group_extract}
g1 <- get_group_result(mg_result, "group1")
class(g1)

estimate_all_total_effects(mg$data_list$group1, g1, method = "ols")
```

`lingam_multi_group_bootstrap()` provides bootstrap stability estimates in
the same joint fashion: every iteration resamples each group independently,
then jointly re-estimates the causal order and per-group adjacency
matrices. It returns a named list of per-group `BootstrapResult` objects, so
the existing bootstrap query functions apply directly per group:

```{r multi_group_bootstrap}
mg_bs <- lingam_multi_group_bootstrap(mg$data_list,
  n_sampling = 20L, reg_method = "ols", seed = 1, verbose = FALSE
)
get_probabilities(mg_bs$group1)
```

Note that `lingam_multi_group_bootstrap()`'s total causal effects are
computed as path-coefficient products over each iteration's adjacency
matrix, not via regression; this matches the upstream Python
implementation but differs from `lingam_direct_bootstrap()`'s
regression-based `estimate_all_total_effects()`.


## Causal Discovery with Missing Data

All algorithms above assume a complete data matrix. When `X` contains missing
values (`NA`), `bootstrap_with_imputation()` combines bootstrap resampling
with multiple imputation: each resample is imputed into several complete
datasets, and a common causal structure is jointly estimated across them with
`lingam_multi_group()` (treating the imputed copies as "groups" that share one
causal order). This is an R port of the Python
`lingam.tools.bootstrap_with_imputation()`.

```{r bootstrap_with_imputation_setup, eval = requireNamespace("mice", quietly = TRUE)}
sample6_na <- generate_lingam_sample_6(n = 1000, seed = 1)
X_na <- sample6_na$data
set.seed(1)
X_na$x5[sample.int(nrow(X_na), size = round(0.1 * nrow(X_na)))] <- NA # MCAR 10% on x5
```

```{r bootstrap_with_imputation_fit, eval = requireNamespace("mice", quietly = TRUE)}
bwi <- bootstrap_with_imputation(X_na,
  n_sampling = 20L, n_repeats = 5L, seed = 42, verbose = FALSE
)
print(bwi)
```

The default imputer is `mice::mice(method = "norm")` (Bayesian linear
regression), the closest standard R equivalent of the upstream Python default
(`IterativeImputer(sample_posterior = TRUE)`); numeric results will not match
the Python implementation. Both the imputer and the causal-discovery step can
be swapped for a custom `function` via the `imputer` and `cd_fit` arguments.

Because each iteration produces `n_repeats` adjacency matrices (one per
imputed dataset), the result's shape differs from `lingam_direct_bootstrap()`.
`as_bootstrap_result()` collapses the `n_repeats` dimension (median or mean)
into a regular `BootstrapResult`, so the existing bootstrap query functions
apply as usual:

```{r bootstrap_with_imputation_convert, eval = requireNamespace("mice", quietly = TRUE)}
bs_na <- as_bootstrap_result(bwi, aggregate = "median")
get_probabilities(bs_na)
```

`get_total_causal_effects()` is not available on this `BootstrapResult`, since
`bootstrap_with_imputation()` never computes total effects.


## Related Articles

- [Method selection guide](method-selection.html) — which method fits your data
- [Direct LiNGAM in depth](direct-lingam.html)
- [Bootstrap and diagnostics](bootstrap-diagnostics.html)
