---
title: "Time Series: VAR-LiNGAM and VARMA-LiNGAM"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5
)
set.seed(0)
```

Direct LiNGAM assumes that observations are **independent and identically
distributed (i.i.d.)**, a requirement that time series data violate. This
article covers the two time-series methods in `lingamr`:

- **VAR-LiNGAM** (`lingam_var()`): stationary time series whose temporal
  dependence is captured by an autoregressive (AR) part.
- **VARMA-LiNGAM** (`lingam_varma()`): time series whose disturbances also
  have a moving-average (MA) part, so past shocks influence the present.

If your data are i.i.d. cross-sectional observations, start with the
[Direct LiNGAM article](direct-lingam.html) instead. For an overview of all
methods, see the [method selection guide](method-selection.html).

```{r setup}
library(lingamr)
```


## VAR-LiNGAM: Causal Discovery in Time Series

**VAR-LiNGAM** (Hyvärinen et al., 2010) handles stationary time series by
first fitting a Vector Autoregression (VAR) model to absorb temporal
autocorrelation, then applying Direct LiNGAM to the VAR residuals to recover
the **instantaneous** causal structure $B_0$.
The model is:

$$X_t = B_0\,X_t + \sum_{k=1}^{p} B_k\,X_{t-k} + e_t$$

where $B_0$ encodes contemporaneous causal effects (strictly acyclic),
$B_1, \ldots, B_p$ encode lagged effects, and $e_t$ are mutually independent
non-Gaussian disturbances.

### Sample Data

`generate_varlingam_sample()` produces a three-variable time series from a
VAR(1)-LiNGAM model.
The instantaneous structure is $x_0 \to x_1 \to x_2$ (coefficients 0.6 and
−0.5), and the only cross-variable lag-1 effect is $x_2(t-1) \to x_0(t)$
(coefficient 0.3).

```{r varlingam_sample}
s <- generate_varlingam_sample(n = 1000, seed = 42)

# True instantaneous coefficient matrix B0  (B0[i, j]: x_j -> x_i)
s$true_B0

# True lag-1 coefficient matrix  (M1[i, j]: x_j(t-1) -> x_i(t), structural)
s$true_M1
```

### Fitting VAR-LiNGAM

Pass the data matrix to `lingam_var()`.
Rows must be in chronological order (earliest first).

```{r lingam_var}
model <- lingam_var(s$data, lags = 1)
model
```

The result object contains `adjacency_matrices`, a three-dimensional array of
shape `[1 + lags, n_features, n_features]`:

- **`[1, , ]` (`"lag0"`):** instantaneous matrix $B_0$.
  `B0[i, j]` is the direct effect of $x_j$ on $x_i$ at the *same* time step.
- **`[k + 1, , ]` (`"lag`*k*`"`):** lagged matrix $B_k$.
  `Bk[i, j]` is the direct structural effect of $x_j(t-k)$ on $x_i(t)$.

Both can be extracted by their dimension label:

```{r varlingam_matrices}
B0 <- model$adjacency_matrices["lag0", , ]
B1 <- model$adjacency_matrices["lag1", , ]

round(B0, 2)  # compare with s$true_B0
round(B1, 2)  # compare with s$true_M1
```

### Lag Order Selection

By default, `lingam_var()` automatically selects the lag order among `1:lags`
using the Bayesian Information Criterion (`criterion = "bic"`).
The alternatives `"aic"`, `"hqic"`, and `"fpe"` are also supported.
To use a fixed lag order without automatic selection, set `criterion = NULL`:

```{r varlingam_lags, eval = FALSE}
# Fix lag order to 2 without IC-based selection
model_lag2 <- lingam_var(s$data, lags = 2, criterion = NULL)
```

### Stationarity Check

VAR-LiNGAM is defined for **stationary** processes.
`check_var_stationarity()` inspects the eigenvalues of the VAR companion matrix:
the process is stationary when all moduli are **strictly less than 1**.

```{r varlingam_stationarity}
check_var_stationarity(model)
```

A `max_modulus` at or above 1 indicates a unit-root or explosive process.
In that case, differencing the series before analysis is recommended.

### Residual Diagnostics

LiNGAM assumes that the error terms $e_t$ are **non-Gaussian**.
`test_varlingam_residual_normality()` tests whether the LiNGAM innovations
$e_t = (I - B_0)\,n_t$ (where $n_t$ are the stored VAR residuals) depart
from normality.
A small p-value (reject $H_0$: Gaussian) supports the model assumption.

```{r varlingam_normality}
test_varlingam_residual_normality(model)
```

`test_varlingam_residual_normality_all()` runs several tests at once and
appends skewness and excess kurtosis columns for a quick overview:

```{r varlingam_normality_all}
test_varlingam_residual_normality_all(model, methods = c("shapiro", "jb"))
```

`plot_varlingam_residual_qq()` draws per-variable normal Q-Q plots.
Deviations from the straight reference line indicate non-Gaussianity.

```{r varlingam_qq, fig.width = 7, fig.height = 3}
plot_varlingam_residual_qq(model)
```

### Total Causal Effects

`estimate_var_total_effect()` estimates the **total** causal effect of one
variable on another, integrating over all direct and mediated paths.
The `from_lag` argument controls the time offset of the cause:
`from_lag = 0` (default) gives the contemporaneous total effect;
`from_lag = 1` gives the one-step-ahead effect of $x_j(t-1)$ on $x_i(t)$.

```{r varlingam_total_effect}
# Total effect x0 -> x2 (contemporaneous)
estimate_var_total_effect(s$data, model, from_index = 1, to_index = 3)

# Total effect x0(t-1) -> x2(t) (one-step-ahead)
estimate_var_total_effect(s$data, model, from_index = 1, to_index = 3, from_lag = 1)
```

Variable indices are **1-based** integers, or column names as character strings.

### Bootstrap

`lingam_var_bootstrap()` quantifies the uncertainty of the estimated structure
by re-running VAR-LiNGAM on **residual bootstrap** samples.
Unlike the Direct LiNGAM bootstrap (which resamples i.i.d. rows),
VAR-LiNGAM holds the fitted values fixed and resamples only the VAR residuals,
preserving the temporal structure of the series.

```{r varlingam_bootstrap}
bs_var <- lingam_var_bootstrap(
  s$data,
  n_sampling = 100L,
  seed       = 42,
  verbose    = FALSE
)
```

`get_var_probabilities()` returns the proportion of bootstrap samples in which
each directed edge was detected.
The column layout mirrors `adjacency_matrices`: the first `n_features` columns
correspond to the instantaneous structure (lag 0), the next `n_features` to
lag 1, and so on.

```{r varlingam_probabilities}
round(get_var_probabilities(bs_var), 2)
```

`get_var_paths()` enumerates all causal paths between two variables found
across bootstrap samples, together with each path's average total effect and
detection probability.

```{r varlingam_paths}
# Paths from x0 to x2 at the same time step (from_lag = 0)
get_var_paths(bs_var, from_index = 1, to_index = 3)
```

```{r varlingam_paths_lag}
# Paths from x0(t-1) to x2(t)  (from_lag = 1)
get_var_paths(bs_var, from_index = 1, to_index = 3, from_lag = 1)
```


## VARMA-LiNGAM: Time Series with Moving-Average Errors

**VARMA-LiNGAM** (Kawahara et al., 2011) extends VAR-LiNGAM with a
moving-average (MA) part: the model is
$x_t = B_0 x_t + \sum_{\tau=1}^{p} \psi_\tau x_{t-\tau} + e_t +
\sum_{\omega=1}^{q} \Omega_\omega e_{t-\omega}$,
so past disturbances can influence the present alongside the lagged
variables. `lingam_varma()` estimates the reduced-form VARMA coefficients by
the deterministic two-stage Hannan-Rissanen procedure (the Python reference
uses state-space maximum likelihood), applies Direct LiNGAM to the residuals,
and returns the AR-side matrices `psis` (with `psis[1, , ]` = B0) and the
MA-side matrices `omegas`.

```{r lingam_varma}
s_varma <- generate_varmalingam_sample(n = 1000, seed = 42)
model_varma <- lingam_varma(s_varma$data, order = c(1, 1))
print(model_varma)
```

`check_varma_stationarity()` checks the AR eigenvalues (stationarity) and
also the MA eigenvalues (invertibility), which Hannan-Rissanen does not
enforce.

```{r varma_stationarity}
check_varma_stationarity(model_varma)
```

### Residual Diagnostics

As in VAR-LiNGAM, the error terms are assumed to be **non-Gaussian**.
`test_varmalingam_residual_normality()` tests whether a VARMA-LiNGAM residual
series departs from normality. By default (`on = "innovations"`) it targets
the LiNGAM innovations $e_t = (I - B_0)\,n_t$, where $n_t$ are the stored
VARMA residuals; set `on = "varma"` to test $n_t$ directly instead. A small
p-value (reject $H_0$: Gaussian) supports the model assumption.

```{r varma_normality}
test_varmalingam_residual_normality(model_varma)
```

`test_varmalingam_residual_normality_all()` runs several tests at once and
appends skewness and excess kurtosis columns for a quick overview:

```{r varma_normality_all}
test_varmalingam_residual_normality_all(model_varma, methods = c("shapiro", "jb"))
```

`plot_varmalingam_residual_qq()` draws per-variable normal Q-Q plots (the same
`on` argument selects the residual series). Deviations from the straight
reference line indicate non-Gaussianity.

```{r varma_qq, fig.width = 7, fig.height = 3}
plot_varmalingam_residual_qq(model_varma)
```

### Total Causal Effects

`estimate_varma_total_effect()` estimates the **total** causal effect of one
variable on another, the VARMA-LiNGAM counterpart of
`estimate_var_total_effect()`: it integrates over both the AR (psi) and MA
(omega) structure. As in VAR-LiNGAM, `from_lag = 0` (default) gives the
contemporaneous total effect and `from_lag = 1` gives the one-step-ahead
effect of $x_j(t-1)$ on $x_i(t)$.

```{r varma_total_effect}
# Total effect x0 -> x2 (contemporaneous)
estimate_varma_total_effect(s_varma$data, model_varma, from_index = 1, to_index = 3)

# Total effect x0(t-1) -> x2(t) (one-step-ahead)
estimate_varma_total_effect(s_varma$data, model_varma, from_index = 1, to_index = 3, from_lag = 1)
```

### Bootstrap

`lingam_varma_bootstrap()`, `get_varma_probabilities()`, and
`get_varma_paths()` mirror their VAR-LiNGAM counterparts, re-running
VARMA-LiNGAM on residual bootstrap samples. In the probability matrix, the
first `1 + p` column blocks are the psi (lag) matrices and the final `q`
blocks are the omega (MA) matrices.

```{r varma_bootstrap}
bs_varma <- lingam_varma_bootstrap(
  s_varma$data,
  n_sampling = 100L,
  order      = c(1, 1),
  criterion  = NULL,
  seed       = 42,
  verbose    = FALSE
)
round(get_varma_probabilities(bs_varma, min_causal_effect = 0.1), 2)
```

### VAR or VARMA?

- Start with **VAR-LiNGAM**: it is simpler, faster, and its lag order can be
  selected automatically by information criteria.
- Switch to **VARMA-LiNGAM** when the VAR residuals remain autocorrelated
  even after increasing the lag order — a sign that the disturbances have an
  MA component that a finite-order VAR can only approximate with many lags.
- Both methods assume stationarity; check with `check_var_stationarity()` /
  `check_varma_stationarity()`, and difference the series when a unit root is
  suspected.


## Related Articles

- [Method selection guide](method-selection.html) — which method fits your data
- [Direct LiNGAM in depth](direct-lingam.html) — the i.i.d. building block
- [Bootstrap and diagnostics](bootstrap-diagnostics.html) — assessing the
  reliability of estimated structures
